Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-04-07 paper.tex
population studies almost finished

We continue by comparing the estimated differential flux from Fornax
to three other clusters in Fig.~\ref{fig:clu_comp}; the close by well
studied Virgo cluster, the X-ray bright Perseus clusters, and the
massive merging Coma cluster. We find high DM induced gamma-ray fluxes
from both the Fornax and Virgo clusters, which highlight them as
promising targets for indirect DM searches. Note, however, that we
have used spatially unresolved upper limits, which is quite optimistic
for the angular extended Virgo cluster. It is quite striking how
constraining the upper limits for the Fornax cluster is compared to
the other clusters, especially in the $1-10$~GeV energy regime where
Fermi-LAT is most sensitive due to the maximized effective area. The
upper limits for Virgo and Perseus are background contaminated by AGN
activity from M87 and NGC 1267, respectively, and do not gain much
from the increased sensitivity. One should also note that the high CRp
induced flux from both Coma and Perseus is of the same magnitude as
the DM flux from LP models, making it difficult for indirect DM
searches in those clusters. Conversely is the Fornax cluster a great
target for indirect DM studies because the relative low gamma-ray flux
from CRps, absents of an active AGN, and high DM gamma-ray
flux. Actually, the Fermi-LAT 3 year data will be able to constrain
the smallest mass of substructures using the DM BM models. We have
also learned that the closest clusters are superior for DM searches
compared to further away clusters, even though the close by clusters
have a mass that is smaller. In fact, we can understand this by
looking at the luminosity-to-mass scaling relations. The gamma-ray
luminosity from the smooth density profile is given by
\begin{equation}
L_{\gamma,\rmn{sm}} \propto \int \dd V \rho(r)^2 \propto \frac{c^3}
{\left[\log\left(1+c\right)-c/(1+c)\right]} \propto \mvir^{0.83}\,,
\end{equation}
and the total DM luminosity including boost factors for the LP and
the BM models are given by
\begin{eqnarray}
L_{\gamma,\rmn{LP}} &=& L_{\gamma,\rmn{smLP}} \B_\rmn{LP} \propto \mvir^{0.73}\qquad
\,\,\,\,\,\,\rmn{and}\\
L_{\gamma,\rmn{BM}} &=& L_{\gamma,\rmn{smBM}} \B_\rmn{BM} \propto  \mvir^{1.06}\,,\qquad
\rmn{where} \\
B_\rmn{LP}&=&B_\sfe B_\sub \propto \mvir^{-1/3}\mvir^{0.23} \propto
 \mvir^{-0.1}\,,\\
B_\rmn{BM}&=&B_\sub \propto \mvir^{0.23}\,.
\end{eqnarray}
Hence increasing the mass with a factor two doubles the luminosity,
but increasing the distance by a factor two decrease the flux with a
factor four. It is also usefull to relate the luminosities to a more
model independent variable as the maximum circular velocity given by
\begin{eqnarray}
V_\rmn{max} &=& V_\rmn{circ}(r_\rmn{max})\qquad\,\,\,\,\,\,\,\,\,\rmn{where}\nonumber\\
V_\rmn{circ}(r)&=& \sqrt{\frac{G_\rmn{N}\,M(<r)}{r}}\qquad\rmn{and}\nonumber\\
\frac{\dd V_\rmn{circ}(r)}{\dd r} &=& 0\quad \Longrightarrow\quad r_\rmn{max}\,.
\end{eqnarray}
We find that $V_\rmn{max}=650\,\rmn{km/s}$ of a large $10^{15}\,\msun$
cluster, and $V_\rmn{max}=420\,\rmn{km/s}$ of a small
$4\times10^{13}\,\msun$ group.

@@

\section{Predictions and observational limits}
Model predictions

\begin{figure}
  \includegraphics[width=0.99\columnwidth]{figures/MLscaling.eps}
  \caption{\it Lum-scaling relations, mass dependence of SD cooling
    have been neglected, actually a factor few difference for
    groups. factor 2 for Fornax, and small difference for large
    clusters}
 \label{fig:lum_mass_scaling}
\end{figure}

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.99\columnwidth]{figures/Fermi.comp.DM.eps}
  \caption{\it Fermi flux upper limits contrasted to the predicted DM
    gamma-ray flux. We show the mean differential flux in the energy
    range $E_\gamma=1-10$~GeV for 32 clusters. The Fermi-LAT upper
    limits are shown with black arrows. The predicted fluxes are
    derived from both a leptophilic DM model that result in inverse
    Compton upscattered CMB photons (red), and the continuum emission
    from the DM $\Kp$ benchmark model (blue). The leptophilic fluxes
    are ruled out by 14 of the clusters, with the strongest
    constraints set by Fornax and M49. We can not constrain the
    benchmark models, although our predictions are approaching the
    upper limits.}
 \label{fig14}
\end{minipage}
\end{figure*}
 difficult to detect DM without a Sommerfeld boost with Cherenkov
 telescopes since the boost from substructures is extended while the
 sensitivity of Cherenkov telescopes drops linearly with radius
 outside the point spread function (CHECK)

we rule out the leptophilic models in present form in 14 clusters,
 and limit the boost from Sommerfeld enhancement in the MW to less
 than 10

mention prime cluster candidates, LP under pressure, Fornax start
being constraining, since only few clusters close to UL stacking will
not increase the DM limits much

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.99\columnwidth]{figures/Fermi.comp.CR.diff.eps}
  \caption{\it Fermi-LAT flux upper limits contrasted to the
    predictions by the semi-analytic Pinzke and Pfrommer model for CRp
    induced gamma-ray emission. We show the mean differential flux in
    the energy range $E_\gamma=1-10$~GeV for 32 clusters. The Fermi
    upper limits are shown with black arrows. The blue boxes show the
    gamma-ray emission from CRp induced $\pi^0$\:s decaying, where the
    upper bound shows the estimates for an optimistic model and the
    lower bound a more conservative model excluding point sources (the
    realistic case is more likely in between, see
    \cite{2010MNRAS.409..449P} for details). Note that there is
    currently no tension between our predictions and upper limits from
    Fermi, although Virgo, Norma, and Coma are close and will in the
    coming years enforce constraints on the hadronic models.}
 \label{fig15}
\end{minipage}
\end{figure*}

@@

@@

\begin{figure*}%[t]
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.99\columnwidth]{figures/Flux.comp.CR.eps}
\caption{\it Comparing the flux from clusters in the extended HIFLUGCS
  catalogue. We show the energy integrated gamma-ray flux induced by
  CRps for the 106 clusters included in the extended HIFLUGCS
  catalogue. The fluxes are calculated within $\rvir$, and are derived
  using a single beta profile for each cluster's gas density profile
  (see text for details). The upper panel show the energy integrated
  flux above 100~MeV (left side) and above 100~GeV (right side), both
  as a function of HIFLUGCS cluster ID. The name of the eight
  brightest clusters are written out. The lower panel show the
  relative difference between the gamma-ray flux predicted by
  mass-luminosity scaling relations compared to the flux computed
  using the semi-analytically model \cite{2010MNRAS.409..449P}. The
  name of the clusters with the largest offset are printed out, which
  correlate with cool core clusters.}
 \label{fig19}
\end{minipage}
\end{figure*}
if CC, then skewed towards large X-ray fluxes; X-ray emission good
proxies for gamma-ray emission --) The dust turns up at about a factor
100 lower than the stars --) The other figure show the spectral shape
of the IC-star with the same spectral shape as the dust, it shows the
same gamma-ray slope as the dust --) The reason why there is only a
factor 10 difference in the break in the gamma-ray emission between
the star and dust is related to the KN. In our CRp paper, the last
figure shows that at $\eg=100\,\tev$, the spectra is suppressed with
about a factor 10 (also depends on $\alpha_\e$). Since the dust
photons have a factor 10 higher energy, I expect this suppression at
about 10~TeV. For star photons the suppression should be at
100~GeV. So the star photons are definitely KN suppressed, which
explains the earlier break and the steeper slope.

>>>>

We continue by comparing the estimated differential flux from Fornax
to three other clusters in Fig.~\ref{fig:clu_comp}; the close by well
studied Virgo cluster, the X-ray bright Perseus clusters, and the
massive merging Coma cluster. We find high DM induced gamma-ray fluxes
from both the Fornax and Virgo clusters, which highlight them as
promising targets for indirect DM searches. Note, however, that we
have used spatially unresolved upper limits, which is quite optimistic
for the angular extended Virgo cluster. It is quite striking how
constraining the upper limits for the Fornax cluster is compared to
the other clusters, especially in the $1-10$~GeV energy regime where
Fermi-LAT is most sensitive due to the maximized effective area. The
upper limits for Virgo and Perseus are background contaminated by AGN
activity from M87 and NGC 1267, respectively, and do not gain much
from the increased sensitivity. One should also note that the high CRp
induced flux from both Coma and Perseus is of the same magnitude as
the DM flux from LP models, making it difficult for indirect DM
searches in those clusters. Conversely is the Fornax cluster a great
target for indirect DM studies because the relative low gamma-ray flux
from CRps, absents of an active AGN, and high DM gamma-ray
flux. Actually, the Fermi-LAT 3 year data will be able to constrain
the smallest mass of substructures using the DM BM models. We have
also learned that the closest clusters are superior for DM searches
compared to further away clusters, even though the close by clusters
have a mass that is smaller. In fact, we can understand this by
looking at the luminosity-to-mass scaling relations. The gamma-ray
luminosity from the smooth density profile is given by
\begin{equation}
L_{\gamma,\rmn{sm}} \propto \int \dd V \rho(r)^2 \propto \frac{c^3}
{\left[\log\left(1+c\right)-c/(1+c)\right]} \propto \mvir^{0.83}\,,
\end{equation}
and the total DM luminosity including boost factors for the LP and
the BM models are given by
\begin{eqnarray}
\label{eq:LP_scaling}
L_{\gamma,\rmn{LP}} &=& L_{\gamma,\rmn{smLP}} \B_\rmn{LP} \propto \mvir^{0.73}\qquad
\,\,\,\,\,\,\rmn{and}\\
\label{eq:BM_scaling}
L_{\gamma,\rmn{BM}} &=& L_{\gamma,\rmn{smBM}} \B_\rmn{BM} \propto  \mvir^{1.06}\,,\qquad
\rmn{where} \\
B_\rmn{LP}&=&B_\sfe B_\sub \propto \mvir^{-1/3}\mvir^{0.23} \propto
 \mvir^{-0.1}\,,\\
B_\rmn{BM}&=&B_\sub \propto \mvir^{0.23}\,.
\end{eqnarray}
Hence increasing the mass with a factor two doubles the luminosity,
but increasing the distance by a factor two decrease the flux with a
factor four. It is also usefull to relate the luminosities to a more
model independent variable as the maximum circular velocity given by
\begin{eqnarray}
V_\rmn{max} &=& V_\rmn{circ}(r_\rmn{max})\qquad\,\,\,\,\,\,\,\,\,\rmn{where}\nonumber\\
V_\rmn{circ}(r)&=& \sqrt{\frac{G_\rmn{N}\,M(<r)}{r}}\qquad\rmn{and}\nonumber\\
\frac{\dd V_\rmn{circ}(r)}{\dd r} &=& 0\quad \Longrightarrow\quad r_\rmn{max}\,.
\end{eqnarray}
We find that $V_\rmn{max}=650\,\rmn{km/s}$ of a large $10^{15}\,\msun$
cluster, and $V_\rmn{max}=420\,\rmn{km/s}$ of a small
$4\times10^{13}\,\msun$ group.

@@

\section{Predictions and observational limits}

In this section, we compute the expected $\gamma$-ray fluxes from DM
annihilation and CR interactions of the brightest clusters of the X-ray flux
complete sample in the local universe. We confront these predictions to upper
limits obtained by Fermi 1.5-year data and conclude on the viability of the
underlying models and perspectives for the next years of Fermi observations.

\begin{figure*}
  \includegraphics[width=0.99\columnwidth]{figures/MLscaling.100M.eps}
  \includegraphics[width=0.99\columnwidth]{figures/MLscaling.10G.eps}
  \caption{\it Scaling relations of the cluster's virial mass $M_{200}$ and the
    $\gamma$-ray luminosity for energies $E_\gamma>100$~MeV (left) and
    $E_\gamma>10$~GeV (right).  Shown are the relations for the CR-induced
    emission that is dominated by pion decay resulting from hadronic CR
    interactions (solid), the leptophilic model of DM annihilations that
    includes the scaling of the substructure and Sommerfeld boost factors with
    custer mass from Equation~(\ref{eq:LP_scaling}) (dash-dotted) as well as the
    benchmark model $K'$ of DM annihilation that includes the scaling of the
    substructure boost with cluster mass of Equation~(\ref{eq:BM_scaling})
    (dashed). Note that the scaling of this particular benchmark model is
    indicative for all (non-Sommerfeld enhanced) DM annihilation models while
    the normalization depends on the particular cross-section and neyralino
    masses.}
\label{fig:lum_mass_scaling}
\end{figure*}

First we focus on $\gamma$-ray flux-cluster mass scaling relations for DM
annihilation and CR induced emission in Fig.~\ref{fig:lum_mass_scaling}.  The
mass scaling of the substructure boost steepens the intrinsically shallower DM
annihilation relation of benchmark models (Equation~(\ref{eq:BM_scaling})) while
this steepening is in fact over-compensated by the inverse mass scaling of the
Sommerfeld boost in leptophilic models that prefers smaller halos with a larger
ratio of $c/v_\mathrm{circ}$ (Equation~(\ref{eq:LP_scaling})). In contrast, the
CR scaling relation is considerably steeper as shown by reference
\cite{2010MNRAS.409..449P}. [ANDERS: add CR scaling formula] This implies
already a general strategy to minimize the CR-induced foreground for DM
annihilation and argues for very nearby groups. We note that these simulations
do not include AGn feedback that is thought to furthermore reduce the baryon
fraction in groups relative to that in clusters \cite{2008ApJ...687L..53P}. The
resulting smaller target density for hadronic CR interactions will furthermore
steepen the $L_\gamma-M$ relation of CR induced emission, making the case for
groups even stronger.

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.99\columnwidth]{figures/Fermi.comp.DM.eps}
  \caption{\it Fermi flux upper limits are contrasted to the predicted DM
    gamma-ray flux. We show the mean differential flux in the energy range
    $E_\gamma=1-10$~GeV for 32 clusters. The Fermi-LAT upper limits are shown
    with black arrows. The predicted fluxes are derived from the dominant
    inverse Compton-scattering of CMB photons in a leptophilic DM model (red),
    and the continuum emission from the DM $\Kp$ benchmark model
    (blue). Assuming a boost factor due to substructures that has a mass
    spectrum extending down to Earth masses, the expected leptophilic fluxes are
    ruled out by lower upper limits in 14 of the clusters, with the strongest
    constraints set by Fornax and M49. We can not yet constrain the benchmark
    models, although our predictions are approaching the upper limits and will
    allow to test these in the upcoming years with better data.}
 \label{fig14}
\end{minipage}
\end{figure*}

Figure~\ref{fig14} compares Fermi upper limits on the $\gamma$-ray flux with
predictions of DM annihilation fluxes in the leptophilic and benchmark models.
To obtain the $\gamma$-ray fluxes, we scale the virial mass of clusters in the
extended HIFLUGCS catalogue \cite{2002ApJ...567..716R} with our scaling
relations of Equations~(\ref{eq:BM_scaling}) and (\ref{eq:LP_scaling}).  We
assume a boost factor due to substructures that has a constant contribution per
decade in substructure mass and has a mass spectrum extending down to Earth
masses. Fermi limits rule out the leptophilic models in their present form with
the mentioned assumptions in 14 clusters, and limit the boost from Sommerfeld
enhancement to less than 4 [?] in Fornax. Assuming universality, this limits the
Sommerfeld boost in the MW to less than 10 (see Fig.~\ref{fig:boost_const}).
The flux level of the Fermi limits are an indirect measure of the ambient
background flux in the $\gamma$-ray sky and/or the presence of strong point
sources such as AGN in the immediate vicinity of the cluster position. Hence, the ratio of
the Fermi upper limits to the predicted annihilation fluxes,
$F_{\mathrm{UL}}/F_{\mathrm{scal}}$, is a good indication of the best cluster
candidates for indirect DM experiments. We identify Fornax, M49, NGC4636, and
Virgo to be prime candidates. Note however, that Virgo extends of $7\deg$ over
the sky which implies a lower surface brightness and lower signal-to-noise.
Fermi limits on individual clusters are expected to improve as $\sqrt{T/1.5
  \mathrm{yr}}$, where $T$ is the total elapsed time of the Fermi mission. We
emphasize that the very inhomogeneous distribution of $F_{\mathrm{UL}}/
F_{\mathrm{scal}}$ makes it unlikely to dramatically improve the limits through
a stacking analysis as the weak signal will dilute the average constraints.
Assuming the projected sensitivity of the future Cherenkov telescope array (CTA)
of [ANDERS: value?], we note that it will be difficult to detect the DM
annihilation signal from clusters without a Sommerfeld boost by Cherenkov
telescopes. This is because the boost from substructures is extended while the
sensitivity of Cherenkov telescopes drops linearly with radius outside the point
spread function.

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.99\columnwidth]{figures/Fermi.comp.CR.diff.eps}
  \caption{\it Fermi-LAT flux upper limits are contrasted to the predictions by
    the semi-analytic Pinzke and Pfrommer model for CRp induced gamma-ray
    emission. We show the mean differential flux in the energy range
    $E_\gamma=1-10$~GeV for 32 clusters. The Fermi upper limits are shown with
    black arrows. The blue boxes show the gamma-ray emission from CRp induced
    $\pi^0$-decay, where the upper bound shows the estimates for an optimistic
    model and the lower bound a more conservative model excluding point sources
    (note that both models assume the maximally allowed acceleration efficiency
    for strong shocks and most likely will have to be corrected down, see
    \cite{2010MNRAS.409..449P} for details). Note that there is currently no
    tension between our predictions and upper limits from Fermi, although Virgo,
    Norma, and Coma are coming close and will enforce constraints on the
    hadronic models in the coming years.}
 \label{fig15}
\end{minipage}
\end{figure*}

@@

\begin{figure*}%[t]
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.99\columnwidth]{figures/Flux.comp.CR.eps}
 \caption{\it Comparison of the gamma-ray flux induced by CRps for the 106
   clusters included in the extended HIFLUGCS catalogue. The fluxes are
   calculated within $\rvir$, and are derived using a single beta profile for
   each cluster's gas density profile as obtained by ROSAT X-ray
   observations. The CR distribution follows the semi-analytically model derived
   by hydrodynamical cosmological cluster simulations \cite{2010MNRAS.409..449P}
   (see text for details). The upper panel shows the energy integrated flux
   above 100~MeV (left side) and above 100~GeV (right side), both as a function
   of HIFLUGCS cluster ID. The eight brightest clusters are labelled. The lower
   panel shows the relative difference between the gamma-ray flux predicted by
   the mass-luminosity scaling relation and to the flux computed using the
   semi-analytical model \cite{2010MNRAS.409..449P}. The name of the clusters
   with the largest offset are printed out, which correlate with cool core
   clusters.}
 \label{fig19}
\end{minipage}
\end{figure*}
 best clusters, stacking unlikely to improve

if CC, then skewed towards large X-ray fluxes; X-ray emission good
proxies for gamma-ray emission --) The dust turns up at about a factor
100 lower than the stars --) The other figure show the spectral shape
of the IC-star with the same spectral shape as the dust, it shows the
same gamma-ray slope as the dust --) The reason why there is only a
factor 10 difference in the break in the gamma-ray emission between
the star and dust is related to the KN. In our CRp paper, the last
figure shows that at $\eg=100\,\tev$, the spectra is suppressed with
about a factor 10 (also depends on $\alpha_\e$). Since the dust
photons have a factor 10 higher energy, I expect this suppression at
about 10~TeV. For star photons the suppression should be at
100~GeV. So the star photons are definitely KN suppressed, which
explains the earlier break and the steeper slope.

@@
